The Design and Implementation of the Synchronous Language \CEU

### 2016-10-01 paper.tex
(>) @safe (>) finalize

From our experience, however, we find that programs often need non-commutative 
concurrent calls.
%
This is the case for logging (e.g., calls to \code{\_printf} in trails in 
parallel) and for redrawing objects in the screen.
%
Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw the 
objects \code{background} and \code{foreground} in trails in parallel.
In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
conflicting pixels, which makes the calls non commutative and prevents the code 
to compile.
%
However, in this case we want to rely on lexical order to always redraw the 
\code{background} object before the \code{foreground} object.
%
Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} to an 
``intimidating name'' \code{\_redraw\_non\_commutative} (ln. 2, to be explicit 
about its effect), and annotate it as safe (ln. 4--5) to make the code compile 
successfully.

@@

.
\end{lstlisting}
\centering\small{\ax \code{\_redraw} \textbf{cannot} be concurrent}
\end{minipage}
%
\begin{minipage}[t]{0.52\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
native do
    #define redraw_non_commutative redraw
end
@safe _redraw_non_commutative with
      _redraw_non_commutative;
par do
    <...> // animate and redraw "background"
        _redraw_non_commutative(background);
with
    <...> // animate and redraw "fore"
        _redraw_non_commutative(foreground);
end
\end{lstlisting}
\centering\small{\bx \code{\_redraw\_non\_commutative} \textbf{can} be concurrent}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Making the non-commutative redrawing calls from \ax to compile in \bx.
\label{lst.redraw}
}
\end{figure}

@@

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources:
%
\begin{itemize}
\item If \CEU \textbf{passes} a pointer to a system call, the pointer 
represents a \textbf{local} resource that requires finalization.
\item If \CEU \textbf{receives} a pointer from a system call return, the 
pointer represents an \textbf{external} resource that requires finalization.
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires 
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
(ln. 5), which is an asynchronous call that transmits the buffer in the 
background.
If the block aborts (ln. 11) before receiving an acknowledge from the 
environment (ln. 9), the local \code{msg} goes out of scope and the external 
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.  
4) returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile without the \code{finalize} construct.

@@

\begin{figure}[b]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
par/or do
   var _buffer_t msg;
   <...> // prepare msg
   finalize
      _send_request(&msg);
   with
      _send_cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
.
\end{lstlisting}
\centering\small{\ax Local resource finalization}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\centering\small{\bx External resource finalization}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{
\CEU enforces the use of finalization to prevent \emph{dangling pointers} for 
local resources and \emph{memory leaks} for external resources.
\label{lst.fin.ceu}
}
\end{figure}

>>>>

From our experience, however, we find that programs often need non-commutative 
concurrent calls.
%
This is the case for logging (e.g., calls to \code{\_printf} in trails in 
parallel) and for redrawing objects in the screen.
%
Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw the 
objects \code{background} and \code{foreground} in trails in parallel.
In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
conflicting pixels, which makes the calls non commutative and prevents the code 
to compile.
%
However, in this case we want to rely on lexical order to always redraw the 
\code{background} object before the \code{foreground} object.
%
Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} as
\code{\_redraw\_non\_commutative} (ln. 2) and annotate it as safe (ln. 4--5) to
make the code compile successfully.
The new function name is explicit about its effect, improving code readability.

@@

.
\end{lstlisting}
\centering\small{\ax \code{\_redraw} \textbf{cannot} be concurrent}
\end{minipage}
%
\begin{minipage}[t]{0.52\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em]
native do
    #define redraw_non_commutative redraw
end
native @safe _redraw_non_commutative with
             _redraw_non_commutative;
par do
    <...> // animate and redraw "background"
        _redraw_non_commutative(background);
with
    <...> // animate and redraw "fore"
        _redraw_non_commutative(foreground);
end
\end{lstlisting}
\centering\small{\bx \code{\_redraw\_non\_commutative} \textbf{can} be concurrent}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Making the non-commutative redrawing calls from \ax to compile in \bx.
\label{lst.redraw}
}
\end{figure}

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}:
%
\begin{itemize}
\item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
pointer represents a \textbf{local} resource (ln. \ax:2) that requires finalization
(ln. \ax:7).
\item If \CEU \textbf{receives} a pointer from a system call return (ln. \bx:4),
the pointer represents an \textbf{external} resource (ln. \bx:2) that requires
finalization (ln. \bx:6).
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires 
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
(ln. 5), which is an asynchronous call that transmits the buffer in the 
background.
If the block aborts (ln. 11) before receiving an acknowledge from the 
environment (ln. 9), the local \code{msg} goes out of scope and the external 
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.  
4) returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile without the \code{finalize} construct.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
par/or do
   var _buffer_t msg;
   <...> // prepare msg
   finalize
      _send_request(&msg);
   with
      _send_cancel(&msg);
   end
   await SEND_ACK;
with
   <...>
end
.
\end{lstlisting}
\centering\small{\ax Local resource finalization}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
par/or do
   var _FILE* f;
   finalize
      f = _fopen(...);
   with
      _fclose(f);
   end
   _fwrite(..., f);
   await A;
   _fwrite(..., f);
with
   <...>
end
\end{lstlisting}
\centering\small{\bx External resource finalization}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{
\CEU enforces the use of finalization to prevent \emph{dangling pointers} for 
local resources and \emph{memory leaks} for external resources.
\label{lst.fin.ceu}
}
\end{figure}

\begin{comment}

@@

@@

\end{comment}
